Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-03-25 WSSSPE2_report/WSSSPE2_review.tex
minor fixes, mostly names, in 3.2

Neil Chue Hong is currently serving as the Director of the Software
Sustainability Institute (SSI), United Kingdom. Hong started with making the
point that world-leading research relies on software, quoting Kersten Kleese Van
Dam of the Pacific Northwestern National Laboratory via the campaign
\href{http://change.org}{change.org}, ``\emph{Today there are very few science
areas left which do not rely on IT and thus software for the majority of the
their research work. More importantly key scientific advances in experimental
and observational science would have been impossible without better software.}''
He also cited Daniel Katz, Sustained Innovation (SI2) Program Director of the
National Science Foundation, ``\emph{Scientific discovery and innovation are
advancing along fundamentally new pathways opened by development of increasingly
sophisticated software. Software is an integral enabler of computation,
experiment and theory, and directly responsible for increased scientific
productivity and enhancement of researchers' capabilities.}''

Hong drew attention to the misconception that we too often disregard hundreds of
thousands of software researchers thinking they are the long tail. We forget
that actually they are the mainstream and we are the elite minority. He
emphasized that software is no longer special; it is prevailing in scientific
research. Survey of researchers from 15 Russell Group universities conducted by
the SSI between Aug--Oct 2014 included a total of 406 respondents covering a
representative range of funders, disciplines, and seniority: 92\% confirmed the
use of research software such as MATLAB, SPSS, Excel, so on and so forth; 89\%
affirmed that it would be impossible or difficult to conduct research without
software. Nevertheless the British research community is just starting to
understand the magnitude of the issue. In the aforementioned SSI survey, 56\% of
researchers developed their own research software; 71\% had no formal
software-development training; 140,000 researchers relied on their own coding
skills; 4\% of jobs advertised in British universities were software related.

According to Hong, costs of software-reliant research in the United Kingdom
included \textsterling 840 million of investment in the financial year
2013--2014, an amount that has risen by 3\% on average over the past four years.
About 30\% of total research investment has been spent on research which relies
on software over the last four financial years. These numbers stemmed from an
analysis of data from $49,650$ grant titles and abstracts published on Gateway
to Research between years 2010 and 2014.

Hong led the audience to ponder the following questions: What are we going to
\emph{do} to help and benefit software researchers and to achieve software
sustainability if~92\% of researchers and WSSSPE care about scientific software?
Who we need to persuade? What are the incentives we need to put in place? He
challenged the workshop participants to change the current deficient practices
in research and academia.

>>>>

Neil Chue Hong is currently serving as the Director of the Software
Sustainability Institute (SSI), United Kingdom. Chue Hong started with making the
point that world-leading research relies on software, quoting Kersten Kleese Van
Dam of the Pacific Northwestern National Laboratory via the campaign
\href{http://change.org}{change.org}, ``\emph{Today there are very few science
areas left which do not rely on IT and thus software for the majority of the
their research work. More importantly key scientific advances in experimental
and observational science would have been impossible without better software.}''
He also cited Daniel Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of the
National Science Foundation, ``\emph{Scientific discovery and innovation are
advancing along fundamentally new pathways opened by development of increasingly
sophisticated software. Software is an integral enabler of computation,
experiment and theory, and directly responsible for increased scientific
productivity and enhancement of researchers' capabilities.}''

Chue Hong drew attention to the fact that we too often disregard hundreds of
thousands of software researchers thinking they are the long tail. We forget
that actually they are the mainstream and we are the elite minority. He
emphasized that software is no longer special; it is both essential to and common in scientific
research. A survey of researchers from 15 Russell Group universities conducted by
the SSI between Aug--Oct 2014, which included a total of 406 respondents covering a
representative range of funders, disciplines, and seniority, reported that 92\% confirmed the
use of research software such as MATLAB, SPSS, Excel, etc. and 89\%
affirmed that it would be impossible or difficult to conduct research without
software. Nevertheless the British research community is just starting to
understand the magnitude of the issue. In the aforementioned SSI survey, 56\% of
researchers developed their own research software, 71\% had no formal
software-development training, 140,000 researchers relied on their own coding
skills, and only 4\% of jobs advertised in British universities were software related.

According to Chue Hong, costs of software-reliant research in the United Kingdom
included \textsterling 840 million of investment in the financial year
2013--2014, and this amount has risen by 3\% on average over the past four years.
About 30\% of total research investment has been spent on research that relies
on software over the last four financial years. These numbers stemmed from an
analysis of data from $49,650$ grant titles and abstracts published on Gateway
to Research between years 2010 and 2014.

Chue Hong led the audience in discussing the following questions: What are we going to
\emph{do} to help and benefit software researchers and to achieve software
sustainability if~92\% of researchers and WSSSPE care about scientific software?
Who do we need to persuade? What are the incentives we need to put in place? He
challenged the workshop participants to change the current deficient practices
in research and academia.
